Tidy Data

### 2014-01-17 tidy-data.tex
Another attempt at describing semantics

A table is a collection of data \textbf{values}, typically either numbers (if quantitative) or strings (if qualitative). A \textbf{variable} is a set of measurements of a single quantifiable property for multiple observational units. An \textbf{observation} is a set of measurements of multiple quantifiable properties for a single observational unit. An observational unit is the smallest constituent atom of the experimental unit. For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at different points in time. In an agricultural experiment, the experimental unit might be the cages assigned different diets, but the observational units are the individual pigs in those cages. In short, while an observational unit represents a conditional property, like blood pressure at time 1 or the weight of pig 2, an experimental unit represents an unconditional property, like blood pressure and weight.

The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} contains three variables that comprise 10 values. The data contains a \code{person} variable with three values (John, Mary, and Jane), a \code{treatment} variable with two values (a and b), and a \code{result} variable with has five or six values depending on how you think of the missing value (-, 16, 3, 2, 11, 1). The layout of Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} makes the number of values clear, but obscures their meaning.

The meaning of the dataset relies on how these values are related to each other. The \code{result} variable was collected over the course of six observations: each \code{person} was observed twice, once for each \code{treatment}. These observations give meaning to the specific values of result. For example, the value 16 is meaningful in the context of the observation: \emph{Jane} scored \emph{16} when she took treatment \code{a}. Table~\ref{tbl:preg-tidy} organizes the dataset around the six observations, making the meaning of the individual values more clear.

>>>>

A table is a collection of data \textbf{values}, usually either numbers (if quantitative) or strings (if qualitative). Values are organised in two ways. Every value belongs to a \textbf{variable} and an \textbf{observation}. A variable collects all values that measure the same underlying attribute (like height, temperature, duration) across units. An observation collects all values measured on the same unit (like a person, or a day, or a race) across attributes.

Table~\ref{tbl:preg-tidy} reorganises the data of Table~\ref{tbl:preg-raw-1} to make the values, variables and obserations more clear. We can see that the dataset contains 18 values representing three variables and six observations. The variables are:

\begin{enumeriate}

\item \code{person}, with three possible values (John, Mary, and Jane),

\item \code{treatment}, with two possible values (a and b), and

\item \code{result}, with five or six values depending on how you think of the missing value (-, 16, 3, 2, 11, 1).

\end{enumerate}

The experimental design tells us more about the structure of the observations. In this experiment, every combination of of \code{person} and \code{treatment} was measured, a completely crossed design.
